---
title: "Data cleaning for final project"
author: "Nathalie Bonnet"
date: 04/22/2026
format: html
---

```{r}
library(tidyverse)
library(janitor)
library(here)
library(lubridate)
library(readxl)
```

# Cleaning nors data

## Reading in raw CSV file and checking that data types appear correctly
```{r}
nors_data <- read.csv(here("data", "raw", "NORS_20260422.csv")) %>% 
  clean_names()

glimpse(nors_data)
```

## View top categories that etiology and setting data fall into

I'm most interested in using state, etiology, setting, and ifsac category to understand if there are patterns in food poisoning cases between states. 

```{r}
nors_data %>%
  count(etiology, sort = TRUE) %>%
  as.data.frame() %>%
  head(50)

nors_data %>%
  count(setting, sort = TRUE) %>%
  as.data.frame() %>%
  head(50)
```

## Etiology and setting mapping
Etiology and setting have a lot of unique values, but can generally be classified into broader categories
```{r}
# Map top pathogens and use keyword matching for the rest
etiology_category_map <- function(etiology) {
  case_when(
    is.na(etiology) | etiology == "" ~ "Unknown",
    
    # Viral
    grepl(
      "Norovirus|Hepatitis A|Rotavirus|Sapovirus|Astrovirus|Adenovirus|Other - Virus|Enterovirus",
      etiology,
      ignore.case = TRUE
    ) ~ "Viral",
    
    # Bacterial
    grepl(
      "Salmonella|Escherichia|Shigella|Clostridium|Campylobacter|Listeria|Legionella|Staphylococcus|Bacillus|Vibrio|Yersinia|Brucella|Cronobacter|Other - Bacterium|Pseudomonas|Streptococcus|Plesiomonas|Leptospira|Enterobacter|Enterococcus|Mycobacterium|Pantoea",
      etiology,
      ignore.case = TRUE
    ) ~ "Bacterial",
    
    # Chemical
    grepl(
      "Copper|Selenium|Zinc|Mercury|Lead|Pesticide|Scombrotoxin|Ciguatoxin|Chemical|Toxin|Histamine|Chlorine|Chloramine|Fluoride|Heavy metal|Sodium Hydroxide|Cleaning|Soap|Nitrite|Nitrate|Arsenic|Chromium|Bromate|Bromine|Phenol|Gasoline|Toluene|Xylene|Ethyl|Morpholine|Herbicide|Trichloroethylene|Hydrochloric|Alkaline Water|Oil|Microcystis|Microcoleus|shellfish poison|Paralytic|Neurotoxic|Amnesic",
      etiology,
      ignore.case = TRUE
    ) ~ "Chemical",
    
    # Parasitic
    grepl(
      "Cryptosporidium|Giardia|Cyclospora|Trichinella|Toxoplasma|Anisakis|Other - Parasite|Avian schistosomes|Entamoeba|Naegleria",
      etiology,
      ignore.case = TRUE
    ) ~ "Parasitic",
    # Others
    grepl("Unknown|unknown", etiology) ~ "Unknown",
    TRUE ~ "Other"
  )
}

# Setting mapping
setting_category_map <- function(setting) {
  case_when(
    is.na(setting) | setting == "" ~ "Unknown",
    grepl("Unknown", setting, ignore.case = TRUE) ~ "Unknown",
    grepl(
      "Restaurant|Caterer|Banquet|Bar|Cafeteria|Food service",
      setting,
      ignore.case = TRUE
    ) ~ "Food Service",
    grepl(
      "Long-term care|nursing home|assisted living|Hospital|Healthcare",
      setting,
      ignore.case = TRUE
    ) ~ "Healthcare",
    grepl(
      "School|college|university|Child daycare|preschool",
      setting,
      ignore.case = TRUE
    ) ~ "School",
    grepl(
      "Private home|residence|Apartment|Condo",
      setting,
      ignore.case = TRUE
    ) ~ "Home",
    grepl(
      "Church|Camp|Fair|Festival|Reception|Wedding|Religious|Event space",
      setting,
      ignore.case = TRUE
    ) ~ "Event",
    grepl(
      "Prison|Correctional|Shelter|group home|transitional",
      setting,
      ignore.case = TRUE
    ) ~ "Institution",
    grepl(
      "Grocery|Bakery|Deli|convenience store",
      setting,
      ignore.case = TRUE
    ) ~ "Store",
    grepl("Hotel|motel|Resort", setting, ignore.case = TRUE) ~ "Hotel/Resort",
    grepl("Farm|dairy|agricultural", setting, ignore.case = TRUE) ~ "Agricultural",
    grepl("Park|Beach|Outdoor|Recreation", setting, ignore.case = TRUE) ~ "Recreation",
    grepl(
      "Office|workplace|Community|municipality|Subdivision|Neighborhood|Other",
      setting,
      ignore.case = TRUE
    ) ~ "Other",
    
    TRUE ~ "Other"
  )
}

# Create new clean dataset to hold updated data categories
nors_clean <- nors_data %>%
  mutate(etiology_category = etiology_category_map(etiology),
         setting_category  = setting_category_map(setting))

# Verify coverage
nors_clean %>% count(etiology_category, sort = TRUE)
nors_clean %>% count(setting_category, sort = TRUE)
```

## Dropping columns that will not be useful in a database

- Too many serotypes to meaningfully analyze across time and state
- info on hospitalizations and info on deaths 
- food truck type has too many unique options
```{r}
nors_clean <- nors_clean %>% 
  select(! c(serotype_or_genotype, info_on_hospitalizations, info_on_deaths))
```



# Cleaning U.S. Census Data to join

```{r}
pov_census <- readxl::read_xls(here("data", "raw", "est23us.xls"), sheet = 1, skip = 3) %>% 
  clean_names()

glimpse(pov_census)

pop_census <- read.csv(here("data", "apportionment.csv")) %>% 
  clean_names()

glimpse(pop_census)

clean_pov <- pov_census %>% 
  select(state_fips_code, postal_code, name, poverty_estimate_all_ages) %>% 
  mutate(poverty_estimate_all_ages = as.numeric(gsub(",", "", poverty_estimate_all_ages)))

clean_pop <- pop_census %>% 
  select(name, geography_type, year, resident_population) %>% 
  mutate(resident_population = as.numeric(gsub(",", "", resident_population)))

census_join <- clean_pov %>% 
  left_join(clean_pop, by = 'name') %>% 
  mutate(poverty_rate = poverty_estimate_all_ages / resident_population * 100)

names(census_join)
names(nors_clean)
```
```{r}
# NORS - select and rename to match schema
nors_clean <- nors_clean %>%
  mutate(
    illnesses = as.integer(illnesses),
    hospitalizations = as.integer(hospitalizations),
    deaths = as.integer(deaths)
  )

# Census - select and rename to match schema
census_clean <- census_join %>%
  filter(geography_type == "State") %>%
  select(state = name, year, population = resident_population, poverty_rate) %>%
  mutate(census_id = row_number()) %>%
  select(census_id, everything())

names(nors_clean)
```


```{r}
# Write to CSV
write.csv(nors_clean, here("data", "processed", "outbreaks.csv"), row.names = FALSE, na = "")
write.csv(census_clean, here("data", "processed", "census.csv"), row.names = FALSE)
```

