---
title: "final-query-viz"
format: html
---
# Setup
Import necessary packages

```{r}
library(tidyverse)
library(ggrepel)
library(geofacet)
library(tilegramsR)
library(sf)
library(plotly)
```

Setup connection from DuckDB database

```{r}
# Establish connection
conn <- DBI::dbConnect(duckdb::duckdb(), 
                       dbdir = "food_poisoning.duckdb")
conn

# List all tables in connection
DBI::dbListTables(conn)

# Load all tables
census <- tbl(conn, "census")
outbreaks <- tbl(conn, "outbreaks")
```

# Clean data

```{r}
# Summarize outbreak severity
state_severity <- outbreaks %>%
  # Join outbreaks with census
  left_join(
    census %>%
      # Group by state in census table
      group_by(state) %>%
      summarise(
        # Calculate average population and poverty percentage
        avg_population = mean(population, na.rm = TRUE),
        avg_poverty_perc = mean(poverty_perc, na.rm = TRUE)
      ),
    # Join on state
    by = "state"
  ) %>%
  # Group by state again in joined table
  group_by(state) %>%
  summarise(
    # Average illnesses, hospitalizations, and death by 100k people in each state
        illnesses_per_100k = sum(illnesses, na.rm = TRUE) / mean(avg_population, na.rm = TRUE) * 100000,
    deaths_per_100k = sum(deaths, na.rm = TRUE) / mean(avg_population, na.rm = TRUE) * 100000,
    hospitalizations_per_100k = sum(hospitalizations, na.rm = TRUE) / mean(avg_population, na.rm = TRUE) * 100000
  ) %>%
  # Gather results in readble table
  collect()
```

# Create state lookup table to use in mapping off names

```{r}
# state abbreviation lookup table to allow geofacet or tilegrams to aggregate tiles by state
abbrev_lookup <- data.frame(
  abbrev = state.abb,
  state_name = state.name
) %>%
  # Use DC abbreviation to override matching issues
  bind_rows(data.frame(abbrev = "DC", state_name = "District Of Columbia"))

# Join lookup to geofacet data to allow mapping
hex_data <- sf_NPR1to1 %>%
  left_join(abbrev_lookup, by = c("state" = "abbrev")) %>%
  # Join state severity table to plot results
  left_join(
    state_severity %>% mutate(state = str_to_title(state)),
    by = c("state_name" = "state")
  )
```

# Visualizations

First I tried making a geofacet map

```{r}
# Use geofacet packge to map state tiles
state_severity %>%
  mutate(state = str_to_title(state)) %>%
  # Join state names from outcomes to geofacet package coordinates
  left_join(us_state_grid2, by = c("state" = "name")) %>%
  # Plot by state tiles
  ggplot(aes(x = col, 
             y = -row, 
             # Fill with illness data from state severity
             fill = illnesses_per_100k)) +
  # Fill background with white
  geom_tile(color = "white", size = 0.5) +
  # Remove labels visually
  geom_text(aes(label = code), color = "white", size = 3, fontface = "bold") +
  # Use viridis color palette
  scale_fill_viridis_c(
    option = "magma",
    # Legend label
    name = "Illnesses\nper 100k",
    # Darker colors are higher values
    direction = -1
  ) +
  # Add labels
  labs(
    title = "Foodborne Illness Rates by State",
    subtitle = "Higher poverty states tend to show greater outbreak burden",
    caption = "Source: CDC NORS & Census Bureau"
  ) +
  # Remove other random labels
  theme_void() +
  # Format title and legend text
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 10, color = "grey40"),
    legend.position = "right"
  )
```

Then I wanted to try another package for fun! I liked the look of hexagonal state shapes. This is essentially the same process but using {tilegramsR}. 

```{r}
# Rinse and repeat first map
ggplot(hex_data) +
  geom_sf(aes(fill = illnesses_per_100k), color = "white", size = 0.5) +
  geom_sf_text(aes(label = state), color = "white", size = 2.5, fontface = "bold") +
  scale_fill_viridis_c(
    option = "viridis",
    name = "Illnesses\nper 100k",
    direction = -1,
      ) +
  labs(
    title = "Foodborne Illness Rates by State",
    caption = "Source: CDC NORS & Census Bureau"
  ) +
  theme_void() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 10, color = "grey40"),
    legend.position = "right"
  )
```
Based on this map, I was interested to see what might be going on with the states that appeared to have higher illness rates. I therefore made flagged state labels to call them out in my final plot. 

In this plot I evaluated illness rates against poverty rates by state. I was surprised to see that there was a negative trend between the two. I expected there to be higher food poisoning illness cases in states with higher poverty rates due to the presence of food deserts and potentially less access to fresh groceries. However, the map and the plot elucidated the opposite. 

This caused me to dig into epidemiology issues, and learned that this is a common calculation issue in epidemiology. States with higher tourism rates or smaller populations artificially look like they have worse health outcomes. This is due to how illnesses are classically calculated against populations, where tourists are not calculated in the state population but still report food poisoning in those states, inflating the reported illnesses in those areas. In states with small populations, even moderate numbers of food poisoning cases result in high calculations of illnesses per people. 

This is apparently a common issue, and one that is inherent in the data, not necessarily solvable with a different calculation. I would try a weighted mean if I were to rerun this analysis. 

```{r}
# Flag notable states to label based on interesting trends from map
notable <- c("New Hampshire", "Nevada", "Rhode Island", "Maine", 
             "Mississippi", "Louisiana", "West Virginia", "California")

# Highlight cases with high death rates or illness rates
state_severity %>%
  mutate(
    state = str_to_title(state),
    highlight = case_when(
      illnesses_per_100k > 2500 ~ "High Illness Rate",
      deaths_per_100k > 7 ~ "High Death Rate",
      TRUE ~ "Other"
    )
  ) %>%
  # Plot on poverty rate and illness rate
  ggplot(aes(x = avg_poverty_perc, y = illnesses_per_100k)) +
  # Plot data points
  geom_point(aes(color = highlight, size = deaths_per_100k), alpha = 0.8) +
  # Plot linear fit to assess if there appears to be a trend
  geom_smooth(method = "lm", se = TRUE, color = "#34B162", linewidth = 0.8) +
  # Add labels to specific data points
  geom_label_repel(
    data = . %>%
    filter(state %in% notable),
    aes(label = state),
    size = 3,
    max.overlaps = 20,
    box.padding = 0.4
  ) +
  # Fill values of interest
  scale_color_manual(
    values = c("High Illness Rate" = "#360855", 
               "High Death Rate" = "#245D7C", 
               "Other" = "#34B162"),
    name = NULL
  ) +
  # Scale size bubbles to not look as dramatic
  scale_size_continuous(name = "Deaths per 100k", range = c(2, 10)) +
  # Scale y axis to adjust plot size
  scale_y_continuous(limits = c(0, 3500)) +
  # Add labels
  labs(
    title = "Poverty Rate vs. Foodborne Illness Burden by State",
    x = "Average Poverty Rate (%)",
    y = "Illnesses per 100k",
    caption = "Source: CDC NORS & Census Bureau"
  ) +
  # Adjust theme
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 10, color = "grey40"),
    legend.position = "bottom"
  )
```
# Close connection!
```{r}
DBI::dbDisconnect(conn, shutdown = TRUE)
```
